---
title: "Example Clinical Trials Tables"
author: "Gabriel Becker and Adrian Waddell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example Clinical Trials Tables}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
suggested_dependent_pkgs <- c("dplyr")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = all(vapply(
    suggested_dependent_pkgs,
    requireNamespace,
    logical(1),
    quietly = TRUE
  ))
)
```

```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```


```{css, echo=FALSE}
.reveal .r code {
    white-space: pre;
}
```


## Introduction

In this vignette we create a

* demographic table
* adverse event table
* response table
* time-to-event analysis table

using the `rtables` layout facility. That is, we demonstrate how the
layout based tabulation framework can specify the structure and
relations that are commonly found when analyzing clinical trials data.

Note that all the data is created using random number generators. All
`ex_*` data which is currently attached to the `rtables` package is
provided by the
[`formatters`](https://pharmaverse.github.io/formatters/)
package and was created using the publicly available
[`random.cdisc.data`](https://insightsengineering.github.io/random.cdisc.data/)
R package.

The packages used in this vignette are:

```{r, message=FALSE}
library(rtables)
library(tibble)
library(dplyr)
```

## Demographic Table

Demographic tables summarize the variables content for different
population subsets (encoded in the columns).

One feature of `analyze()` that we have not introduced in the previous
vignette is that the analysis function `afun` can specify multiple
rows with the `in_rows()` function:

```{r}
ADSL <- ex_adsl # Example ADSL dataset

lyt <- basic_table() |>
  split_cols_by("ARM") |>
  analyze(vars = "AGE", afun = function(x) {
    in_rows(
      "Mean (sd)" = rcell(c(mean(x), sd(x)), format = "xx.xx (xx.xx)"),
      "Range" = rcell(range(x), format = "xx.xx - xx.xx")
    )
  })

tbl <- build_table(lyt, ADSL)
tbl
```

Multiple variables can be analyzed in one `analyze()` call:

```{r}
lyt2 <- basic_table() |>
  split_cols_by("ARM") |>
  analyze(vars = c("AGE", "BMRKR1"), afun = function(x) {
    in_rows(
      "Mean (sd)" = rcell(c(mean(x), sd(x)), format = "xx.xx (xx.xx)"),
      "Range" = rcell(range(x), format = "xx.xx - xx.xx")
    )
  })

tbl2 <- build_table(lyt2, ADSL)
tbl2
```

Hence, if `afun` can process different data vector types
(i.e. variables selected from the data) then we are fairly close to a
standard demographic table. Here is a function that either creates a
count table or some number summary if the argument `x` is a factor or
numeric, respectively:


```{r}
s_summary <- function(x) {
  if (is.numeric(x)) {
    in_rows(
      "n" = rcell(sum(!is.na(x)), format = "xx"),
      "Mean (sd)" = rcell(c(mean(x, na.rm = TRUE), sd(x, na.rm = TRUE)), format = "xx.xx (xx.xx)"),
      "IQR" = rcell(IQR(x, na.rm = TRUE), format = "xx.xx"),
      "min - max" = rcell(range(x, na.rm = TRUE), format = "xx.xx - xx.xx")
    )
  } else if (is.factor(x)) {
    vs <- as.list(table(x))
    do.call(in_rows, lapply(vs, rcell, format = "xx"))
  } else {
    stop("type not supported")
  }
}
```

Note we use `rcell` to wrap the results in order to add formatting
instructions for `rtables`. We can use `s_summary` outside the context
of tabulation:

```{r}
s_summary(ADSL$AGE)
```

and

```{r}
s_summary(ADSL$SEX)
```

We can now create a commonly used variant of the demographic table:

```{r}
summary_lyt <- basic_table() |>
  split_cols_by(var = "ARM") |>
  analyze(c("AGE", "SEX"), afun = s_summary)

summary_tbl <- build_table(summary_lyt, ADSL)
summary_tbl
```

Note that `analyze()` can also be called multiple times in sequence:

```{r}
summary_lyt2 <- basic_table() |>
  split_cols_by(var = "ARM") |>
  analyze("AGE", s_summary) |>
  analyze("SEX", s_summary)

summary_tbl2 <- build_table(summary_lyt2, ADSL)
summary_tbl2
```

which leads to the table identical to `summary_tbl`:

```{r}
identical(summary_tbl, summary_tbl2)
```

```{r, echo=FALSE}
stopifnot(identical(summary_tbl, summary_tbl2))
```

In clinical trials analyses the number of patients per column is often
referred to as `N` (rather than the overall population which outside
of clinical trials is commonly referred to as `N`). Column `N`s are
added by setting the `show_colcounts` argument in `basic_table()` to
`TRUE`:

```{r}
summary_lyt3 <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ARMCD") |>
  analyze(c("AGE", "SEX"), s_summary)

summary_tbl3 <- build_table(summary_lyt3, ADSL)
summary_tbl3
```

### Variations on the Demographic Table

We will now show a couple of variations of the demographic table that
we developed above. These variations are in structure and not in
analysis, hence they don't require a modification to the `s_summary`
function.


We will start with a standard table analyzing the variables `AGE` and
`BMRKR2` variables:

```{r}
lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ARM") |>
  analyze(c("AGE", "BMRKR2"), s_summary)

tbl <- build_table(lyt, ADSL)
tbl
```

Assume we would like to have this analysis carried out per gender
encoded in the row space:

```{r, warning=FALSE}
lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ARM") |>
  split_rows_by("SEX") |>
  analyze(c("AGE", "BMRKR2"), s_summary)

tbl <- build_table(lyt, ADSL)
tbl
```

We will now subset `ADSL` to include only males and females in the
analysis in order to reduce the number of rows in the table:

```{r, warning=FALSE}
ADSL_M_F <- filter(ADSL, SEX %in% c("M", "F"))

lyt2 <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ARM") |>
  split_rows_by("SEX") |>
  analyze(c("AGE", "BMRKR2"), s_summary)

tbl2 <- build_table(lyt2, ADSL_M_F)
tbl2
```

Note that the `UNDIFFERENTIATED` and `U` levels still show up in the
table. This is because tabulation respects the factor levels and level
order, exactly as the `split` and `table` function do. If empty levels
should be dropped then `rtables` needs to know that at splitting time
via the `split_fun` argument in `split_rows_by()`. There are a number
of predefined functions. For this example `drop_split_levels()` is
required to drop the empty levels at splitting time. Splitting is a
big topic and will be eventually addressed in a specific package
vignette.


```{r}
lyt3 <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ARM") |>
  split_rows_by("SEX", split_fun = drop_split_levels, child_labels = "visible") |>
  analyze(c("AGE", "BMRKR2"), s_summary)

tbl3 <- build_table(lyt3, ADSL_M_F)
tbl3
```

In the table above the labels `M` and `F` are not very
descriptive. You can add the full labels as follows:

```{r}
ADSL_M_F_l <- ADSL_M_F |>
  mutate(lbl_sex = case_when(
    SEX == "M" ~ "Male",
    SEX == "F" ~ "Female",
    SEX == "U" ~ "Unknown",
    SEX == "UNDIFFERENTIATED" ~ "Undifferentiated"
  ))

lyt4 <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ARM") |>
  split_rows_by("SEX", labels_var = "lbl_sex", split_fun = drop_split_levels, child_labels = "visible") |>
  analyze(c("AGE", "BMRKR2"), s_summary)

tbl4 <- build_table(lyt4, ADSL_M_F_l)
tbl4
```

For the next table variation we only stratify by gender for the `AGE`
analysis. To do this the `nested` argument has to be set to `FALSE` in
`analyze()` call:

```{r}
lyt5 <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ARM") |>
  split_rows_by("SEX", labels_var = "lbl_sex", split_fun = drop_split_levels, child_labels = "visible") |>
  analyze("AGE", s_summary, show_labels = "visible") |>
  analyze("BMRKR2", s_summary, nested = FALSE, show_labels = "visible")

tbl5 <- build_table(lyt5, ADSL_M_F_l)
tbl5
```

Once we split the rows into groups (`Male` and `Female` here) one
might want to summarize groups: usually by showing count and column
percentages. This is especially important if we have missing data. For
example, if we create the above table but add missing data to the
`AGE` variable:

```{r}
insert_NAs <- function(x) {
  x[sample(c(TRUE, FALSE), length(x), TRUE, prob = c(0.2, 0.8))] <- NA
  x
}

set.seed(1)
ADSL_NA <- ADSL_M_F_l |>
  mutate(AGE = insert_NAs(AGE))

lyt6 <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ARM") |>
  split_rows_by(
    "SEX",
    labels_var = "lbl_sex",
    split_fun = drop_split_levels,
    child_labels = "visible"
  ) |>
  analyze("AGE", s_summary) |>
  analyze("BMRKR2", s_summary, nested = FALSE, show_labels = "visible")

tbl6 <- build_table(lyt6, filter(ADSL_NA, SEX %in% c("M", "F")))
tbl6
```

Here it is not easy to see how many females and males there are in
each arm as `n` represents the number of non-missing data elements in
the variables. Groups within rows that are defined by splitting can be
summarized with `summarize_row_groups()`, for example:

```{r}
lyt7 <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ARM") |>
  split_rows_by("SEX", labels_var = "lbl_sex", split_fun = drop_split_levels) |>
  summarize_row_groups() |>
  analyze("AGE", s_summary) |>
  analyze("BMRKR2", afun = s_summary, nested = FALSE, show_labels = "visible")

tbl7 <- build_table(lyt7, filter(ADSL_NA, SEX %in% c("M", "F")))
tbl7
```

There are a couple of things to note here:

* Group summaries produce "content" rows. Visually, it's impossible to
  distinguish data rows from content rows. Their difference is
  justified (and it's an important design decision) because when we
  paginate tables the content rows are by default repeated if a group
  gets divided via pagination.
* Conceptually the content rows summarize the patient population which
  is analyzed and hence are often the count & group percentages
  (default behavior of `summarize_row_groups()`).

We can recreate this default behavior (count percentage) by defining a
`cfun` for illustrative purposes here as it results in the same table
as above:

```{r}
lyt8 <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ARM") |>
  split_rows_by("SEX", labels_var = "lbl_sex", split_fun = drop_split_levels) |>
  summarize_row_groups(cfun = function(df, labelstr, .N_col, ...) {
    in_rows(
      rcell(nrow(df) * c(1, 1 / .N_col), format = "xx (xx.xx%)"),
      .labels = labelstr
    )
  }) |>
  analyze("AGE", s_summary) |>
  analyze("BEP01FL", afun = s_summary, nested = FALSE, show_labels = "visible")

tbl8 <- build_table(lyt8, filter(ADSL_NA, SEX %in% c("M", "F")))
tbl8
```

Note that `cfun`, like `afun` (which is used in `analyze()`), can
operate on either variables, passed via the `x` argument, or
`data.frame`s or `tibble`s, which are passed via the `df` argument
(`afun` can optionally request `df` too). Unlike `afun`, `cfun` must
accept `labelstr` as the second argument which gives the default group
label (factor level from splitting) and hence it could be modified:

```{r}
lyt9 <- basic_table() |>
  split_cols_by(var = "ARM") |>
  split_rows_by("SEX", labels_var = "lbl_sex", split_fun = drop_split_levels, child_labels = "hidden") |>
  summarize_row_groups(cfun = function(df, labelstr, .N_col, ...) {
    in_rows(
      rcell(nrow(df) * c(1, 1 / .N_col), format = "xx (xx.xx%)"),
      .labels = paste0(labelstr, ": count (perc.)")
    )
  }) |>
  analyze("AGE", s_summary) |>
  analyze("BEP01FL", s_summary, nested = FALSE, show_labels = "visible")

tbl9 <- build_table(lyt9, filter(ADSL_NA, SEX %in% c("M", "F")))
tbl9
```

### Using Layouts

Layouts have a couple of advantages over tabulating the tables
directly:

* the creation of layouts requires the analyst to describe the problem
  in an abstract way
   - i.e. they separate the analyses description from the actual data
* referencing variable names happens via strings (no non-standard
  evaluation (NSE) is needed, though this is arguably either a feature
  or a shortcoming)
* layouts can be reused

Here is an example that demonstrates the reusability of layouts:

```{r}
adsl_lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARM") |>
  analyze(c("AGE", "SEX"), afun = s_summary)

adsl_lyt
```

We can now build a table for `ADSL`

```{r}
adsl_tbl <- build_table(adsl_lyt, ADSL)
adsl_tbl
```

or for all patients that are older than 18:

```{r}
adsl_f_tbl <- build_table(lyt, ADSL |> filter(AGE > 18))
adsl_f_tbl
```


## Adverse Events

There are a number of different adverse event tables. We will now
present two tables that show adverse events by ID and then by grade
and by ID.

This time we won't use the `ADAE` dataset from
[`random.cdisc.data`](https://insightsengineering.github.io/random.cdisc.data/)
but rather generate a dataset on the fly (see [Adrian's 2016 Phuse
paper](https://github.com/waddella/phuse2016_adverse_events)):

```{r}
set.seed(1)

lookup <- tribble(
  ~AEDECOD,                          ~AEBODSYS,                                         ~AETOXGR,
  "HEADACHE",                        "NERVOUS SYSTEM DISORDERS",                        "5",
  "BACK PAIN",                       "MUSCULOSKELETAL AND CONNECTIVE TISSUE DISORDERS", "2",
  "GINGIVAL BLEEDING",               "GASTROINTESTINAL DISORDERS",                      "1",
  "HYPOTENSION",                     "VASCULAR DISORDERS",                              "3",
  "FAECES SOFT",                     "GASTROINTESTINAL DISORDERS",                      "2",
  "ABDOMINAL DISCOMFORT",            "GASTROINTESTINAL DISORDERS",                      "1",
  "DIARRHEA",                        "GASTROINTESTINAL DISORDERS",                      "1",
  "ABDOMINAL FULLNESS DUE TO GAS",   "GASTROINTESTINAL DISORDERS",                      "1",
  "NAUSEA (INTERMITTENT)",           "GASTROINTESTINAL DISORDERS",                      "2",
  "WEAKNESS",                        "MUSCULOSKELETAL AND CONNECTIVE TISSUE DISORDERS", "3",
  "ORTHOSTATIC HYPOTENSION",         "VASCULAR DISORDERS",                              "4"
)

normalize <- function(x) x / sum(x)
weightsA <- normalize(c(0.1, dlnorm(seq(0, 5, length.out = 25), meanlog = 3)))
weightsB <- normalize(c(0.2, dlnorm(seq(0, 5, length.out = 25))))

N_pop <- 300
ADSL2 <- data.frame(
  USUBJID = seq(1, N_pop, by = 1),
  ARM = sample(c("ARM A", "ARM B"), N_pop, TRUE),
  SEX = sample(c("F", "M"), N_pop, TRUE),
  AGE = 20 + rbinom(N_pop, size = 40, prob = 0.7)
)

l.adae <- mapply(
  ADSL2$USUBJID,
  ADSL2$ARM,
  ADSL2$SEX,
  ADSL2$AGE,
  FUN = function(id, arm, sex, age) {
    n_ae <- sample(0:25, 1, prob = if (arm == "ARM A") weightsA else weightsB)
    i <- sample(seq_len(nrow(lookup)), size = n_ae, replace = TRUE, prob = c(6, rep(1, 10)) / 16)
    lookup[i, ] |>
      mutate(
        AESEQ = seq_len(n()),
        USUBJID = id, ARM = arm, SEX = sex, AGE = age
      )
  },
  SIMPLIFY = FALSE
)

ADAE2 <- do.call(rbind, l.adae)
ADAE2 <- ADAE2 |>
  mutate(
    ARM = factor(ARM, levels = c("ARM A", "ARM B")),
    AEDECOD = as.factor(AEDECOD),
    AEBODSYS = as.factor(AEBODSYS),
    AETOXGR = factor(AETOXGR, levels = as.character(1:5))
  ) |>
  select(USUBJID, ARM, AGE, SEX, AESEQ, AEDECOD, AEBODSYS, AETOXGR)

ADAE2
```

### Adverse Events By ID

We start by defining an events summary function:

```{r}
s_events_patients <- function(x, labelstr, .N_col) {
  in_rows(
    "Total number of patients with at least one event" =
      rcell(length(unique(x)) * c(1, 1 / .N_col), format = "xx (xx.xx%)"),
    "Total number of events" = rcell(length(x), format = "xx")
  )
}
```

So, for a population of `5` patients where

* one patient has 2 `AE`s
* one patient has 1 `AE`
* three patients have no `AE`s

we would get the following summary:

```{r}
s_events_patients(x = c("id 1", "id 1", "id 2"), .N_col = 5)
```

The `.N_col` argument is a special keyword argument by which
`build_table()` passes the population size for each respective
column. For a list of keyword arguments for the functions passed to
`afun` in `analyze()`, refer to the documentation with `?analyze`.

We now use the `s_events_patients` summary function in a tabulation:

```{r}
adae_lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARM") |>
  analyze("USUBJID", s_events_patients)

adae_tbl <- build_table(adae_lyt, ADAE2)
adae_tbl
```

Note that the column `N`s are wrong as by default they are set to the
number of rows per group (i.e. number of `AE`s per arm here). This also
affects the percentages. For this table we are interested in the
number of patients per column/arm which is usually taken from `ADSL`
(var `ADSL2` here).

`rtables` handles this by allowing us to override how the column
counts are computed. We can specify an `alt_counts_df` in
`build_table()`. When we do this, `rtables` calculates the column counts
by applying the same column faceting to `alt_counts_df` as it does to
the primary data during tabulation:


```{r}
adae_adsl_tbl <- build_table(adae_lyt, ADAE2, alt_counts_df = ADSL2)
adae_adsl_tbl
```

Alternatively, if the desired column counts are already calculated,
they can be specified directly via the `col_counts` argument to
`build_table()`, though specifying an `alt_counts_df` is the preferred
mechanism (the number of rows will be used, but no duplicate checking!!!).


We next calculate this information per system organ class:

```{r}
adae_soc_lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARM") |>
  analyze("USUBJID", s_events_patients) |>
  split_rows_by("AEBODSYS", child_labels = "visible", nested = FALSE) |>
  summarize_row_groups("USUBJID", cfun = s_events_patients)

adae_soc_tbl <- build_table(adae_soc_lyt, ADAE2, alt_counts_df = ADSL2)
adae_soc_tbl
```

We now have to add a count table of `AEDECOD` for each `AEBODSYS`. The
default `analyze()` behavior for a factor is to create the count table
per level (using `rtab_inner`):

```{r}
adae_soc_lyt2 <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARM") |>
  split_rows_by("AEBODSYS", child_labels = "visible", indent_mod = 1) |>
  summarize_row_groups("USUBJID", cfun = s_events_patients) |>
  analyze("AEDECOD", indent_mod = -1)

adae_soc_tbl2 <- build_table(adae_soc_lyt2, ADAE2, alt_counts_df = ADSL2)
adae_soc_tbl2
```

The `indent_mod` argument enables relative indenting changes if the
tree structure of the table does not result in the desired indentation
by default.

This table so far is however not the usual adverse event table as it
counts the total number of events and not the number of subjects for
one or more events for a particular term. To get the correct table we
need to write a custom analysis function:

```{r}
table_count_once_per_id <- function(df, termvar = "AEDECOD", idvar = "USUBJID") {
  x <- df[[termvar]]
  id <- df[[idvar]]

  counts <- table(x[!duplicated(id)])

  in_rows(
    .list = as.vector(counts),
    .labels = names(counts)
  )
}

table_count_once_per_id(ADAE2)
```

So the desired `AE` table is:

```{r}
adae_soc_lyt3 <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARM") |>
  split_rows_by("AEBODSYS", child_labels = "visible", indent_mod = 1) |>
  summarize_row_groups("USUBJID", cfun = s_events_patients) |>
  analyze("AEDECOD", afun = table_count_once_per_id, show_labels = "hidden", indent_mod = -1)

adae_soc_tbl3 <- build_table(adae_soc_lyt3, ADAE2, alt_counts_df = ADSL2)
adae_soc_tbl3
```

Note that we are missing the overall summary in the first two
rows. This can be added with an initial `analyze()` call.

```{r}
adae_soc_lyt4 <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARM") |>
  analyze("USUBJID", afun = s_events_patients) |>
  split_rows_by("AEBODSYS", child_labels = "visible", indent_mod = 1, section_div = "") |>
  summarize_row_groups("USUBJID", cfun = s_events_patients) |>
  analyze("AEDECOD", table_count_once_per_id, show_labels = "hidden", indent_mod = -1)

adae_soc_tbl4 <- build_table(adae_soc_lyt4, ADAE2, alt_counts_df = ADSL2)
adae_soc_tbl4
```

Finally, if we wanted to prune the 0 count rows we can do that with the `trim_rows()` function:

```{r}
trim_rows(adae_soc_tbl4)
```

Pruning is a larger topic with a [separate `rtables` package
vignette](https://pharmaverse.github.io/rtables/latest-tag/articles/sorting_pruning.html).

### Adverse Events By ID and By Grade

The adverse events table by ID and by grade shows how many patients
had at least one adverse event per grade for different subsets of the
data (e.g. defined by system organ class).

For this table we do not show the zero count grades. Note that we add
the "overall" groups with a custom split function.

```{r}
table_count_grade_once_per_id <- function(df,
                                          labelstr = "",
                                          gradevar = "AETOXGR",
                                          idvar = "USUBJID",
                                          grade_levels = NULL) {
  id <- df[[idvar]]
  grade <- df[[gradevar]]

  if (!is.null(grade_levels)) {
    stopifnot(all(grade %in% grade_levels))
    grade <- factor(grade, levels = grade_levels)
  }

  id_sel <- !duplicated(id)

  in_rows(
    "--Any Grade--" = sum(id_sel),
    .list = as.list(table(grade[id_sel]))
  )
}

table_count_grade_once_per_id(ex_adae, grade_levels = 1:5)
```

All of the layouting concepts needed to create this table have already
been introduced so far:

```{r}
adae_grade_lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARM") |>
  analyze(
    "AETOXGR",
    afun = table_count_grade_once_per_id,
    extra_args = list(grade_levels = 1:5),
    var_labels = "- Any adverse events -",
    show_labels = "visible"
  ) |>
  split_rows_by("AEBODSYS", child_labels = "visible", indent_mod = 1) |>
  summarize_row_groups(cfun = table_count_grade_once_per_id, format = "xx", indent_mod = 1) |>
  split_rows_by("AEDECOD", child_labels = "visible", indent_mod = -2) |>
  analyze(
    "AETOXGR",
    afun = table_count_grade_once_per_id,
    extra_args = list(grade_levels = 1:5),
    show_labels = "hidden"
  )

adae_grade_tbl <- build_table(adae_grade_lyt, ADAE2, alt_counts_df = ADSL2)
adae_grade_tbl
```

## Response Table

The response table that we will create here is composed of 3 parts:

1. Binary response table
2. Unstratified analysis comparison vs. control group
3. Multinomial response table

Let's start with the first part which is fairly simple to derive:

```{r}
ADRS_BESRSPI <- ex_adrs |>
  filter(PARAMCD == "BESRSPI") |>
  mutate(
    rsp = factor(AVALC %in% c("CR", "PR"), levels = c(TRUE, FALSE), labels = c("Responders", "Non-Responders")),
    is_rsp = (rsp == "Responders")
  )

s_proportion <- function(x, .N_col) {
  in_rows(
    .list = lapply(
      as.list(table(x)),
      function(xi) rcell(xi * c(1, 1 / .N_col), format = "xx.xx (xx.xx%)")
    )
  )
}

rsp_lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARMCD", ref_group = "ARM A") |>
  analyze("rsp", s_proportion, show_labels = "hidden")

rsp_tbl <- build_table(rsp_lyt, ADRS_BESRSPI)
rsp_tbl
```


Note that we did set the `ref_group` argument in `split_cols_by()`
which for the current table had no effect as we only use the cell data
for the responder and non-responder counts. The `ref_group` argument
is needed for the part 2 and 3 of the table.

We will now look the implementation of part 2: unstratified analysis
comparison vs. control group. Let's start with the analysis function:

```{r}
s_unstrat_resp <- function(x, .ref_group, .in_ref_col) {
  if (.in_ref_col) {
    return(in_rows(
      "Difference in Response Rates (%)" = rcell(numeric(0)),
      "95% CI (Wald, with correction)" = rcell(numeric(0)),
      "p-value (Chi-Squared Test)" = rcell(numeric(0)),
      "Odds Ratio (95% CI)" = rcell(numeric(0))
    ))
  }

  fit <- stats::prop.test(
    x = c(sum(x), sum(.ref_group)),
    n = c(length(x), length(.ref_group)),
    correct = FALSE
  )

  fit_glm <- stats::glm(
    formula = rsp ~ group,
    data = data.frame(
      rsp = c(.ref_group, x),
      group = factor(rep(c("ref", "x"), times = c(length(.ref_group), length(x))), levels = c("ref", "x"))
    ),
    family = binomial(link = "logit")
  )

  in_rows(
    "Difference in Response Rates (%)" = non_ref_rcell(
      (mean(x) - mean(.ref_group)) * 100,
      .in_ref_col,
      format = "xx.xx"
    ),
    "95% CI (Wald, with correction)" = non_ref_rcell(
      fit$conf.int * 100,
      .in_ref_col,
      format = "(xx.xx, xx.xx)"
    ),
    "p-value (Chi-Squared Test)" = non_ref_rcell(
      fit$p.value,
      .in_ref_col,
      format = "x.xxxx | (<0.0001)"
    ),
    "Odds Ratio (95% CI)" = non_ref_rcell(
      c(
        exp(stats::coef(fit_glm)[-1]),
        exp(stats::confint.default(fit_glm, level = .95)[-1, , drop = FALSE])
      ),
      .in_ref_col,
      format = "xx.xx (xx.xx - xx.xx)"
    )
  )
}

s_unstrat_resp(
  x = ADRS_BESRSPI |> filter(ARM == "A: Drug X") |> pull(is_rsp),
  .ref_group = ADRS_BESRSPI |> filter(ARM == "B: Placebo") |> pull(is_rsp),
  .in_ref_col = FALSE
)
```

Hence we can now add the next vignette to the table:

```{r}
rsp_lyt2 <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARMCD", ref_group = "ARM A") |>
  analyze("rsp", s_proportion, show_labels = "hidden") |>
  analyze(
    "is_rsp", s_unstrat_resp,
    show_labels = "visible",
    var_labels = "Unstratified Response Analysis"
  )

rsp_tbl2 <- build_table(rsp_lyt2, ADRS_BESRSPI)
rsp_tbl2
```

Next we will add part 3: the multinomial response table. To do so, we
are adding a row-split by response level, and then doing the same
thing as we did for the binary response table above.

```{r}
s_prop <- function(df, .N_col) {
  in_rows(
    "95% CI (Wald, with correction)" = rcell(binom.test(nrow(df), .N_col)$conf.int * 100, format = "(xx.xx, xx.xx)")
  )
}

s_prop(
  df = ADRS_BESRSPI |> filter(ARM == "A: Drug X", AVALC == "CR"),
  .N_col = sum(ADRS_BESRSPI$ARM == "A: Drug X")
)
```

We can now create the final response table with all three parts:

```{r}
rsp_lyt3 <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARMCD", ref_group = "ARM A") |>
  analyze("rsp", s_proportion, show_labels = "hidden") |>
  analyze(
    "is_rsp", s_unstrat_resp,
    show_labels = "visible", var_labels = "Unstratified Response Analysis"
  ) |>
  split_rows_by(
    var = "AVALC",
    split_fun = reorder_split_levels(neworder = c("CR", "PR", "SD", "PD", "NE"), drlevels = TRUE),
    nested = FALSE
  ) |>
  summarize_row_groups() |>
  analyze("AVALC", afun = s_prop)

rsp_tbl3 <- build_table(rsp_lyt3, ADRS_BESRSPI)
rsp_tbl3
```

In the case that we wanted to rename the levels of `AVALC` and remove
the CI for `NE` we could do that as follows:

```{r}
rsp_label <- function(x) {
  rsp_full_label <- c(
    CR = "Complete Response (CR)",
    PR = "Partial Response (PR)",
    SD = "Stable Disease (SD)",
    `NON CR/PD` = "Non-CR or Non-PD (NON CR/PD)",
    PD = "Progressive Disease (PD)",
    NE = "Not Evaluable (NE)",
    Missing = "Missing",
    `NE/Missing` = "Missing or unevaluable"
  )
  stopifnot(all(x %in% names(rsp_full_label)))
  rsp_full_label[x]
}


rsp_lyt4 <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARMCD", ref_group = "ARM A") |>
  analyze("rsp", s_proportion, show_labels = "hidden") |>
  analyze(
    "is_rsp", s_unstrat_resp,
    show_labels = "visible", var_labels = "Unstratified Response Analysis"
  ) |>
  split_rows_by(
    var = "AVALC",
    split_fun = keep_split_levels(c("CR", "PR", "SD", "PD"), reorder = TRUE),
    nested = FALSE
  ) |>
  summarize_row_groups(cfun = function(df, labelstr, .N_col) {
    in_rows(nrow(df) * c(1, 1 / .N_col), .formats = "xx (xx.xx%)", .labels = rsp_label(labelstr))
  }) |>
  analyze("AVALC", afun = s_prop) |>
  analyze("AVALC", afun = function(x, .N_col) {
    in_rows(rcell(sum(x == "NE") * c(1, 1 / .N_col), format = "xx.xx (xx.xx%)"), .labels = rsp_label("NE"))
  }, nested = FALSE)

rsp_tbl4 <- build_table(rsp_lyt4, ADRS_BESRSPI)
rsp_tbl4
```

Note that the table is missing the rows gaps to make it more
readable. The row spacing feature is on the `rtables` roadmap and will
be implemented in future.

## Time to Event Analysis Table

The time to event analysis table that will be constructed consists of
four parts:

1. Overall subject counts
2. Censored subjects summary
3. Cox proportional-hazards analysis
4. Time-to-event analysis

The table is constructed by sequential use of the `analyze()`
function, with four custom analysis functions corresponding to each of
the four parts listed above. In addition the table includes
referential footnotes relevant to the table contents. The table will
be faceted column-wise by arm.

First we will start by loading the necessary packages and preparing
the data to be used in the construction of this table.

```{r}
library(survival)

adtte <- ex_adaette |>
  dplyr::filter(PARAMCD == "AETTE2", SAFFL == "Y")

# Add censoring to data for example
adtte[adtte$AVAL > 1.0, ] <- adtte[adtte$AVAL > 1.0, ] |> mutate(AVAL = 1.0, CNSR = 1)

adtte2 <- adtte |>
  mutate(CNSDTDSC = ifelse(CNSDTDSC == "", "__none__", CNSDTDSC))
```

The `adtte` dataset will be used in preparing the models while the
`adtte2` dataset handles missing values in the "Censor Date
Description" column and will be used to produce the final table. We
add censoring into the data for example purposes.

Next we create a basic analysis function, `a_count_subjs` which prints
the overall unique subject counts and percentages within the data.

```{r}
a_count_subjs <- function(x, .N_col) {
  in_rows(
    "Subjects with Adverse Events n (%)" = rcell(length(unique(x)) * c(1, 1 / .N_col), format = "xx (xx.xx%)")
  )
}
```

Then an analysis function is created to generate the counts of
censored subjects for each level of a factor variable in the
dataset. In this case the `cnsr_counter` function will be applied with
the `CNSDTDSC` variable which contains a censor date description for
each censored subject.

```{r}
cnsr_counter <- function(df, .var, .N_col) {
  x <- df[!duplicated(df$USUBJID), .var]
  x <- x[x != "__none__"]
  lapply(table(x), function(xi) rcell(xi * c(1, 1 / .N_col), format = "xx (xx.xx%)"))
}
```

This function generates counts and fractions of unique subjects
corresponding to each factor level, excluding missing values
(uncensored patients).

A Cox proportional-hazards (Cox P-H) analysis is generated next with a
third custom analysis function, `a_cph`. Prior to creating the
analysis function, the Cox P-H model is fit to our data using the
`coxph()` and `Surv()` functions from the `survival` package. Then
this model is used as input to the `a_cph` analysis function which
returns hazard ratios, 95% confidence intervals, and p-values
comparing against the reference group - in this case the leftmost
column.

```{r}
cph <- coxph(Surv(AVAL, CNSR == 0) ~ ACTARM + STRATA1, ties = "exact", data = adtte)

a_cph <- function(df, .var, .in_ref_col, .ref_full, full_cox_fit) {
  if (.in_ref_col) {
    ret <- replicate(3, list(rcell(NULL)))
  } else {
    curtrt <- df[[.var]][1]
    coefs <- coef(full_cox_fit)
    sel_pos <- grep(curtrt, names(coefs), fixed = TRUE)
    hrval <- exp(coefs[sel_pos])
    sdf <- survdiff(Surv(AVAL, CNSR == 0) ~ ACTARM + STRATA1, data = rbind(df, .ref_full))
    pval <- (1 - pchisq(sdf$chisq, length(sdf$n) - 1)) / 2
    ci_val <- exp(unlist(confint(full_cox_fit)[sel_pos, ]))
    ret <- list(
      rcell(hrval, format = "xx.x"),
      rcell(ci_val, format = "(xx.x, xx.x)"),
      rcell(pval, format = "x.xxxx | (<0.0001)")
    )
  }
  in_rows(
    .list = ret,
    .names = c("Hazard ratio", "95% confidence interval", "p-value (one-sided stratified log rank)")
  )
}
```

The fourth and final analysis function, `a_tte`, generates a time to
first adverse event table with three rows corresponding to Median, 95%
Confidence Interval, and Min Max respectively. First a survival table
is constructed from the summary table of a survival model using the
`survfit()` and `Surv()` functions from the `survival` package. This
table is then given as input to `a_tte` which produces the table of
time to first adverse event consisting of the previously mentioned
summary statistics.

```{r}
surv_tbl <- as.data.frame(
  summary(survfit(Surv(AVAL, CNSR == 0) ~ ACTARM, data = adtte, conf.type = "log-log"))$table
) |>
  tibble::rownames_to_column("ACTARM") |>
  dplyr::mutate(
    ACTARM = factor(gsub("ACTARM=", "", ACTARM), levels = levels(adtte$ACTARM)),
    ind = FALSE
  )

a_tte <- function(df, .var, kp_table) {
  ind <- grep(df[[.var]][1], row.names(kp_table), fixed = TRUE)
  minmax <- range(df[["AVAL"]])
  mm_val_str <- format_value(minmax, format = "xx.x, xx.x")
  rowfn <- list()
  if (all(df$CNSR[df$AVAL == minmax[2]])) {
    mm_val_str <- paste0(mm_val_str, "*")
    rowfn <- "* indicates censoring"
  }
  in_rows(
    Median = kp_table[ind, "median", drop = TRUE],
    "95% confidence interval" = unlist(kp_table[ind, c("0.95LCL", "0.95UCL")]),
    "Min Max" = mm_val_str,
    .formats = c("xx.xx", "xx.xx - xx.xx", "xx"),
    .row_footnotes = list(NULL, NULL, rowfn)
  )
}
```

Additionally, the `a_tte` function creates a referential footnote
within the table to indicate where censoring occurred in the data.

Now we are able to use these four analysis functions to build our time
to event analysis table.

```{r}
lyt <- basic_table(show_colcounts = TRUE) |>
  ## Column faceting
  split_cols_by("ARM", ref_group = "A: Drug X") |>
  ## Overall count
  analyze("USUBJID", a_count_subjs, show_labels = "hidden") |>
  ## Censored subjects summary
  analyze("CNSDTDSC", cnsr_counter, var_labels = "Censored Subjects", show_labels = "visible") |>
  ## Cox P-H analysis
  analyze("ARM", a_cph, extra_args = list(full_cox_fit = cph), show_labels = "hidden") |>
  ## Time-to-event analysis
  analyze(
    "ARM", a_tte,
    var_labels = "Time to first adverse event", show_labels = "visible",
    extra_args = list(kp_table = surv_tbl),
    table_names = "kapmeier"
  )

tbl_tte <- build_table(lyt, adtte2)
```

We set the `show_colcounts` argument of `basic_table()` to `TRUE` to
first print the total subject counts for each column. Next we use
`split_cols_by()` to split the table into three columns corresponding
to the three different levels of `ARM`, and specify that the first
arm, `"A: Drug X"` should act as the reference group to be compared
against - this reference group is used for the Cox P-H analysis. Then
we call `analyze()` sequentially using each of the four custom
analysis functions as argument `afun` and specifying additional
arguments where necessary. Then we use `build_table()` to construct
our `rtable` using the `adtte2` dataset.

Finally, we annotate the table using the `fnotes_at_path()` function
to specify that product-limit estimates are used to calculate the
statistics listed under the "Time to first adverse event" heading
within the table. The referential footnote created earlier in the
time-to-event analysis function (`a_tte`) is also displayed.

```{r}
fnotes_at_path(
  tbl_tte,
  c("ma_USUBJID_CNSDTDSC_ARM_kapmeier", "kapmeier")
) <- "Product-limit (Kaplan-Meier) estimates."

tbl_tte
```
