---
title: "Introductory rtables - Facet And Analysis Nesting"
author: "Gabriel Becker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introductory rtables - Facet And Analysis Nesting}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

# Introduction


`rtables` models data-summarizing tables as *faceted data
visualizations*, analogous to a `ggplot2` plot using `facet_grid` or a
`lattice` plot conditioned on multiple factors.

We saw in the previous section that we use:

- `split_cols_by` to declare *columns*,
- `split_rows_by` to declare *groups of individual rows*,
- `summarize_row_groups` to declare *marginal summary rows* for groups
  of individual rows, and
- `analyze` to declare (sets of) *individual rows*.

Combining a single call each to `split_cols_by`, `split_rows_by` and
`analyze` creates a rectangular table, while adding
`summarize_row_groups` after the `split_rows_by` adds marginal summary
rows for each group.

Often we need tables with more complex structure, whether it is
multiple top-level sections of the table; tables which analyze
multiple variables simultaneously; nested faceting in row structure,
column structure, or both; or combinations of all three of these.

We achieve all of these by leveraging *nesting* of layout instructions.

# Nesting

*Nesting* is how we talk about *where* a layout instruction fits with
respect to the existing state of the layout. We say an instruction is
*nested within* a preceding faceting instruction (`split_rows_by` or
`split_cols_by`) if the new instruction *should be applied separately
within each facet generated during tabulation from the previous
instruction. This is analogous to what we see with `facet_*` in
`ggplot2` when we give multiple variables for a single faceting
dimension.

By default, each layout instruction is nested within the directly
preceding layout instruction - if any - in its dimension (row or
column), with a couple caveats we discuss later. We see this default
behavior below:

```{r}
library(rtables)
lyt <- basic_table() |>
  split_cols_by("ARM") |>
  split_cols_by("STRATA1") |>
  split_rows_by("SEX") |>
  split_rows_by("BMRKR2") |>
  analyze("AGE")
build_table(lyt, ex_adsl)
```

When `analyze` instructions are 'nested within' another `analyze`, the
analyses are bundled into a 'multi-analysis' parent structure. This
parent structure as a whole, then, has the nesting behavior that an
single `analyze` call would have in its place.

```{r}
lyt2 <- basic_table() |>
  split_cols_by("ARM") |>
  split_cols_by("STRATA1") |>
  split_rows_by("SEX") |>
  split_rows_by("BMRKR2") |>
  analyze("AGE") |>
  analyze("BMRKR1")
head(build_table(lyt2, ex_adsl), 32)
```


By default:

- `analyze` calls nest within the most recently preceding
  `split_rows_by` or instruction
  - multiple `analyze` calls that nest within the 
