---
title: "Introductory rtables - Facet And Analysis Nesting"
author: "Gabriel Becker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introductory rtables - Facet And Analysis Nesting}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

# Introduction


`rtables` models data-summarizing tables as *faceted data
visualizations*, analogous to a `ggplot2` plot using `facet_grid` or a
`lattice` plot conditioned on multiple factors.

We saw in the previous section that we use:

- `split_cols_by` to declare *columns*,
- `split_rows_by` to declare *groups of individual rows*,
- `summarize_row_groups` to declare *marginal summary rows* for groups
  of individual rows, and
- `analyze` to declare (sets of) *individual rows*.

Combining a single call each to `split_cols_by`, `split_rows_by` and
`analyze` creates a rectangular table, while adding
`summarize_row_groups` after the `split_rows_by` adds marginal summary
rows for each group.

Often we need tables with more complex structure, whether it is
multiple top-level sections of the table; tables which analyze
multiple variables simultaneously; nested faceting in row structure,
column structure, or both; or combinations of all three of these.

We achieve all of these by leveraging *nesting* of layout instructions.

# Nesting

*Nesting* is how we talk about *where* a layout instruction fits with
respect to the existing state of the layout. We say an instruction is
*nested within* a preceding faceting instruction (`split_rows_by` or
`split_cols_by`) if the new instruction *should be applied separately
within each facet generated during tabulation from the previous
instruction. This is analogous to what we see with `facet_*` in
`ggplot2` when we give multiple variables for a single faceting
dimension.

By default, each layout instruction is nested within the directly
preceding layout instruction - if any - in its dimension (row or
column), with a couple caveats we discuss later. We see this default
behavior below:

```{r}
library(rtables)
lyt <- basic_table() |>
  split_cols_by("ARM") |>
  split_cols_by("STRATA1") |>
  split_rows_by("SEX") |>
  split_rows_by("BMRKR2") |>
  analyze("AGE")
table_structure(build_table(lyt, ex_adsl))
```

When `analyze` instructions are 'nested within' another `analyze`, the
analyses are bundled into a 'multi-analysis' parent structure. This
parent structure as a whole, then, has the nesting behavior that a
single `analyze` call would have in its place.

```{r}
lyt2 <- basic_table() |>
  split_cols_by("ARM") |>
  split_cols_by("STRATA1") |>
  split_rows_by("SEX") |>
  split_rows_by("BMRKR2") |>
  analyze("AGE") |>
  analyze("BMRKR1")
table_structure(build_table(lyt2, ex_adsl))
```

By default:

- `analyze` calls nest within the most recently preceding
  `split_rows_by` or instruction
  - multiple `analyze` calls that nest within the 

## Creating 'Multi-Section' Tables With `nested = FALSE`

We often want to create tables with rows grouped into two or more
logical or analytical sections. For example we might want to analyze
`AGE` overall, then separately by `SEX` and by `RACE`. We can do this by:

1. `analyze`ing `AGE`, then
2. splitting by `SEX` and `analyze`ing `AGE`, and finally
3. splitting by `RACE` and `analyze`ing `AGE` again.

We will start each section after the first with `nested = FALSE` to
delineate it from the previous portion of the layout.

NOTE: while we will do it explicitly for illustration purposes, any
`split_rows_by` layout instruction that follows an `analyze` is
defaults to `nested = FALSE`.

Thus we can create our table with the code below:

Note: we set a top level section divider to make our different
sections concrete; section dividers will be covered in a later part of
this guide and can be taken as is for now.

```{r}
trim_adsl <- subset(ex_adsl, RACE %in% levels(ex_adsl$RACE)[1:3] & SEX %in% c("F", "M"))
trim_adsl$RACE <- factor(trim_adsl$RACE)
trim_adsl$SEX <- factor(trim_adsl$SEX)

nice_mean <- function(x) {
  in_rows("Average Age" = mean(x), .formats = list("Average Age" = "xx.x"))
}

lyt3 <- basic_table(top_level_section_div = "-") |>
  split_cols_by("ARM") |>
  analyze("AGE", afun = nice_mean) |>
  split_rows_by("SEX", nested = FALSE) |>
  analyze("AGE", afun = nice_mean) |>
  split_rows_by("RACE", nested = FALSE) |>
  analyze("AGE", afun = nice_mean)

tbl3 <- build_table(lyt3, trim_adsl)
tbl3
```

We see three clear top-level 'sections' of our table in row-space, as
desired.  Contrast this with our result without `nested = FALSE` (and
with the first two `analyze` calls replaced with
`summarize_row_groups`:


```{r}
nice_mean_cfun <- function(x, labelstr) {
  lbl <- paste0(labelstr, " (Ave. Age)")
  in_rows(mean(x), .labels = lbl, .formats = "xx.x")
}

lyt3b <- basic_table(top_level_section_div = "-") |>
  split_cols_by("ARM") |>
  summarize_row_groups("AGE", cfun = nice_mean_cfun) |>
  split_rows_by("SEX") |>
  summarize_row_groups("AGE", cfun = nice_mean_cfun) |>
  split_rows_by("RACE") |>
  analyze("AGE", afun = nice_mean)

tbl3b <- build_table(lyt3b, trim_adsl)
head(tbl3b)
```

Here the faceting on `RACE` occurs *nested within* the faceting on
`SEX`, whereas above it occurs *alongside* it.

### A Brief Note On Removing Sections

Sometimes we will receive a template script that does more than we
want, but is close to meeting our needs. For example, imagine we
wanted the above table (non-nested version) but only wanted the
overall and race portions, removing the age within gender analysis.

We do this by simply identifying all of the layout instructions
corresponding to that portion of the table and removing them. Our top
level section dividers can help us reason about this, and can be added
to the template if they were not there originally.

In our case, the instructions for our section to remove are the
`split_rows_by("SEX", nested = FALSE)`, and directly following
`analyze("AGE")` calls. By starting with our code above and removing
those, we would get our desired table:


```{r}
lyt3c <- basic_table(top_level_section_div = "-") |>
  split_cols_by("ARM") |>
  analyze("AGE", afun = nice_mean) |>
  ## split_rows_by("SEX", nested = FALSE) |>
  ## analyze("AGE", afun = nice_mean) |>
  split_rows_by("RACE", nested = FALSE) |>
  analyze("AGE", afun = nice_mean)

tbl3c <- build_table(lyt3c, trim_adsl)
tbl3c
```

When performing this kind of layout pruning in the wild, it is
important to remember that `split_rows_by` calls that follow `analyze`
calls default to `nested = FALSE`, even if that is not made explicit
in the template script you are starting from.

It is also important to not remove all `analyze` call(s) nested within
any series of row faceting (`split_rows_by*` calls), as this will
result in an degenerate (invalidly structured) table which could have
undefined behavior when passed to some other aspects of the `rtables`
and `formatters` APIs.

## Intermediate Nesting

As of `rtables` `0.7.0`, we can declare *intermediate* nesting, rather
than simply full -- the previous and now default behavior when `nested
= TRUE` -- and no -- the `nested = FALSE` behavior -- nesting.

We do this via the new `at_sibling` parameter the `split_rows_by*` and
`analyze*` families of layout functions now accept.

This allows us to, for example, split by strata, further split by
gender, analyze strata x gender, then analyze the strata as a whole,
like so:


```{r}
lyt4 <- basic_table() |>
  split_cols_by("ARM") |>
  split_rows_by("STRATA1") |>
  split_rows_by("SEX", split_fun = keep_split_levels(c("F", "M"))) |>
  analyze("AGE") |>
  analyze("BMRKR2", at_sibling = "SEX")

tbl <- build_table(lyt4, ex_adsl)

lyt4b <- basic_table() |>
  split_cols_by("ARM") |>
  split_rows_by("STRATA1") |>
  split_rows_by("SEX", split_fun = keep_split_levels(c("F", "M"))) |>
  analyze("AGE") |>
  analyze("BMRKR2", at_sibling = "STRATA1")

tbl4b <- build_table(lyt4b, ex_adsl)

```



```{r}
lyt5 <- basic_table() |>
  split_cols_by("ARM") |>
  split_rows_by("STRATA1") |>
  split_rows_by("SEX", split_fun = keep_split_levels(c("F", "M"))) |>
  analyze("AGE") |>
  split_rows_by("RACE", split_fun = keep_split_levels(levels(ex_adsl$RACE)[1:2]), at_sibling = "SEX") |>
  analyze("AGE")

tbl <- build_table(lyt5, ex_adsl)
```
