---
title: "Table Hierarchy"
author: "Abinaya Yogasekaram"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Disclaimer

This article is intended for use by developers only and will contain low-level explanations of the topics covered. For user-friendly vignettes, please see the [Articles](https://pharmaverse.github.io/rtables/latest-tag/articles/index.html) page on the `rtables` website.

Any code or prose which appears in the version of this article on the `main` branch of the repository may reflect a specific state of things that can be more or less recent. This guide describes very important aspects of table hierarchy that are unlikely to change. Regardless, we invite the reader to keep in mind that the current repository code may have drifted from the following material in this document, and it is always the best practice to read the code directly on `main`.

Please keep in mind that `rtables` is still under active development, and it has seen the efforts of multiple contributors across different years. Therefore, there may be legacy mechanisms and ongoing transformations that could look different in the future.

## Introduction

The scope of this vignette is to understand the structure of `rtable` objects, class hierarchy with an exploration of tree structures as S4 objects. Exploring table structure enables a better understanding of `rtables` concepts such as split machinery, tabulation, pagination and export. More details from the user's perspective of table structure can be found in the relevant vignettes.

`isS4`
`getclass` - for class structure


## Process and Methods

We invite developers to use the provided examples to interactively explore the `rtables` hierarchy. The most helpful command is `getClass` for a list of the slots associated with a class, in addition to related classes and their relative distances.

## Representation of Information before generation


## Table Representation
`PredataAxisLayout` class is used to define the data subset instructions for tabulation. 2 sub-classes (one for each axis): `PredataColLayout`, `PredataRowLayout`

## Slots, Parent-Child Relationships

## Content (summary row groups)

Splits are core functionality for `rtables` as tabulation and calculations are often required on subsets of the data.

## Split Machinery
```{r, message=FALSE}
library(rtables)
getClass("TreePos")
```

`TreePos` class contains split information as a list of the splits, split label values, and the subsets of the data that are generated by the split.

`AllSplit`
`RootSplit`
`MultiVarSplit`
`VarStaticCutSplit`
`CumulativeCutSplit`
`VarDynCutSplit`
`CompoundSplit`
`VarLevWBaselineSplit`


The highest level of the table hierarchy belong to `TableTree`. The code below identifies the slots associated with with this class.
```{r}
getClass("TableTree")
```

As an S4 object, the slots can be accessed using `@` (similar to the use of `$` for list objects).
You'll notice there are classes that fall under "Extends". The classes contained here have a relationship to the `TableTree` object and are "virtual" classes. To avoid the repetition of slots and carrying the same data (set of slots for example) that multiple classes may need, `rtables` extensively uses virtual classes. A virtual class cannot be instantiated, the purpose is for other classes to inherit information from it.


```{r}
lyt <- basic_table(title = "big title") |>
  split_rows_by("SEX", page_by = TRUE) |>
  analyze("AGE")

tt <- build_table(lyt, DM)

# Though we don't recommend using str for studying rtable objects,
# we do find it useful in this instance to visualize the parent/child relationships.
str(tt, max.level = 2)
```

## Tree Paths

Root to Leaves, are vectors of vectors
Tables are tree, nodes in the tree can have summaries associated with them. Tables are trees because of the nested structure. There is also the benefit of keeping and repeating necessary information when trying to paginate a table.

Children of `ElementaryTables` are row objects. `TableTree` can have children that are either row objects or other table objects.


#### TODO:
Create Tree Diagram showing class hierarchy.
