---
title: "Tabulation"
author: "Davide Garolini"
date: '`r Sys.Date()`'
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Disclaimer

This article is intended for use by developers only and will contain low-level explanations of the topics covered. For user-friendly vignettes, please see the [Articles](https://pharmaverse.github.io/rtables/latest-tag/articles/index.html) page on the `rtables` website.

Any code or prose which appears in the version of this article on the `main` branch of the repository may reflect a specific state of things that can be more or less recent. This guide describes very important aspects of the tabulation process that are unlikely to change. Regardless, we invite the reader to keep in mind that the current repository code may have drifted from the following material in this document, and it is always the best practice to read the code directly on `main`.

Please keep in mind that `rtables` is still under active development, and it has seen the efforts of multiple contributors across different years. Therefore, there may be legacy mechanisms and ongoing transformations that could look different in the future.

Being that this a working document that may be subjected to both deprecation and updates, we keep `xxx` comments to indicate placeholders for warnings and to-do's that need further work.

## Introduction

Tabulation in `rtables` is a process that takes a pre-defined layout and applies it to data. The layout object, with all of its splits and `analyze`s, can be applied to different data to produce valid tables. This process happens principally within the `tt_dotabulation.R` file and the user-facing function `build_table` that resides in it. We will occasionally use functions and methods that are present in other files, like `colby_construction.R` or `make_subset_expr.R`. We assume the reader is already familiar with the documentation for `build_table`. We suggest reading the [Split Machinery article](https://pharmaverse.github.io/rtables/latest-tag/articles/dev-guide/dg_split_machinery.html) prior to this one, as it is instrumental in understanding how the layout object, which is essentially built out of splits, is tabulated when data is supplied.

## Tabulation

We enter into `build_table` using `debugonce` to see how it works.

```{r, eval=FALSE}
# rtables 0.6.2
library(rtables)
debugonce(build_table)

# A very simple layout
lyt <- basic_table() |>
  split_rows_by("STRATA1") |>
  split_rows_by("SEX", split_fun = drop_split_levels) |>
  split_cols_by("ARM") |>
  analyze("BMRKR1")

# lyt must be a PreDataTableLayouts object
is(lyt, "PreDataTableLayouts")

lyt |> build_table(DM)
```

Now let's look within our `build_table` call. After the initial check that the layout is a pre-data table layout, it checks if the column layout is defined (`clayout` accessor), i.e. it does not have any column split. If that is the case, a `All obs` column is added automatically with all observations. After this, there are a couple of defensive programming calls that do checks and transformations as we finally have the data. These can be divided into two categories: those that mainly concern the layout, which are defined as generics, and those that concern the data, which are instead a function as they are not dependent on the layout class. Indeed, the layout is structured and can be divided into `clayout` and `rlayout` (column and row layout). The first one is used to create `cinfo`, which is the general object and container of the column splits and information. The second one contains the obligatory all data split, i.e. the root split (accessible with `root_spl`), and the row splits' vectors which are iterative splits in the row space. In the following, we consider the initial checks and defensive programming.

```{r, eval=FALSE}
## do checks and defensive programming now that we have the data
lyt <- fix_dyncuts(lyt, df) # Create the splits that depends on data
lyt <- set_def_child_ord(lyt, df) # With the data I set the same order for all splits
lyt <- fix_analyze_vis(lyt) # Checks if the analyze last split should be visible
# If there is only one you will not get the variable name, otherwise you get it if you
# have multivar. Default is NA. You can do it now only because you are sure to
# have the whole layout.
df <- fix_split_vars(lyt, df, char_ok = is.null(col_counts))
# checks if split vars are present

lyt[] # preserve names - warning if names longer, repeats the name value if only one
lyt@.Data # might not preserve the names # it works only when it is another class that inherits from lists
# We suggest doing extensive testing about these behaviors in order to do choose the appropriate one
```

Along with the various checks and defensive programming, we find `PreDataAxisLayout` which is a virtual class that both row and column layouts inherit from. Virtual classes are handy for group classes that need to share things like labels or functions that need to be applicable to their relative classes. See more information about the `rtables` class hierarchy in the dedicated article [here](https://pharmaverse.github.io/rtables/latest-tag/articles/dev-guide/dg_table_hierarchy.html).

Now, we continue with `build_table`. After the checks, we notice `TreePos()` which is a constructor for an object that retains a representation of the tree position along with split values and labels. This is mainly used by `create_colinfo`, which we enter now with `debugonce(create_colinfo)`. This function creates the object that represents the column splits and everything else that may be related to the columns. In particular, the column counts are calculated in this function. The parameter inputs are as follows:

```{r, eval=FALSE}
cinfo <- create_colinfo(
  lyt, # Main layout with col split info
  df, # df used for splits and col counts if no alt_counts_df is present
  rtpos, # TreePos (does not change out of this function)
  counts = col_counts, # If we want to overwrite the calculations with df/alt_counts_df
  alt_counts_df = alt_counts_df, # alternative data for col counts
  total = col_total, # calculated from build_table inputs (nrow of df or alt_counts_df)
  topleft # topleft information added into build_table
)
```

`create_colinfo` is in `make_subset_expr.R`. Here, we see that if `topleft` is present in `build_table`, it will override the one in `lyt`. Entering `create_colinfo`, we will see the following calls:

```{r, eval=FALSE}
clayout <- clayout(lyt) # Extracts column split and info

if (is.null(topleft)) {
  topleft <- top_left(lyt) # If top_left is not present in build_table, it is taken from lyt
}

ctree <- coltree(clayout, df = df, rtpos = rtpos) # Main constructor of LayoutColTree
# The above is referenced as generic and principally represented as
# setMethod("coltree", "PreDataColLayout", (located in `tree_accessor.R`).
# This is a call that restructures information from clayout, df, and rtpos
# to get a more compact column tree layout. Part of this design is related
# to past implementations.

cexprs <- make_col_subsets(ctree, df) # extracts expressions in a compact fashion.
# WARNING: removing NAs at this step is automatic. This should
# be coupled with a warning for NAs in the split (xxx)

colextras <- col_extra_args(ctree) # retrieves extra_args from the tree. It may not be used
```

Next in the function is the determination of the column counts. Currently, this happens only at the leaf level, but it can certainly be calculated independently for all levels (this is an open issue in `rtables`, i.e. how to print other levels' totals). Precedence for column counts may be not documented ("xxx todo"). The main use case is when you are analyzing a participation-level dataset, with multiple records per subject, and you would like to retain the total numbers of subjects per column, often taken from a subject-level dataset, to use as column counts. Originally, counts were only able to be added as a vector, but it is often the case that users would like the possibility to use `alt_counts_df`. The `cinfo` object (`InstantiatedColumnInfo`) is created with all the above information.

If we continue inside `build_table`, we see `.make_ctab` used to make a root split. This is a general procedure that generates the initial root split as a content row. `ctab` is applied to this content row, which is a row that contains only a label. From `?summarize_row_groups`, you know that this is how `rtables` defines label rows, i.e. as content rows. `.make_ctab` is very similar to the function that actual creates the table rows, `.make_tablerows`. Note that this function uses `parent_cfun` and `.make_caller` to retrieve the content function inserted in above levels. Here we split the structural handling of the table object and the row-creation engine, which are divided by a `.make_tablerows` call. If you search the package, you will find that this function is only called twice, once in `.make_ctab` and once in `.make_analyzed_tab`. These two are the final elements of the table construction: the creation of rows.

Going back to `build_table`, you will see that the row layout is actually a list of split vectors. The fundamental line, `kids <- lapply(seq_along(rlyt), function(i) {`, allows us to appreciate this. Going forward we see how `recursive_applysplit` is applied to each split vector. It may be worthwhile to check what this vector looks like in our test case.

```{r, eval=FALSE}
# rtables 0.6.2
# A very simple layout
lyt <- basic_table() |>
  split_rows_by("STRATA1") |>
  split_rows_by("SEX", split_fun = drop_split_levels) |>
  split_cols_by("ARM") |>
  analyze("BMRKR1")

rlyt <- rtables:::rlayout(lyt)
str(rlyt, max.level = 2)
```

```c
Formal class 'PreDataRowLayout' [package "rtables"] with 2 slots
  ..@ .Data     :List of 2 # rlyt is a rtables object (PreDataRowLayout) that is also a list!
  ..@ root_split:Formal class 'RootSplit' [package "rtables"] with 17 slots # another object!
  # If you do summarize_row_groups before anything you act on the root split. We need this to
  # have a place for the content that is valid for the whole table.

str(rtables:::root_spl(rlyt), max.level = 2) # it is still a split

str(rlyt[[1]], max.level = 3) # still a rtables object (SplitVector) that is a list
Formal class 'SplitVector' [package "rtables"] with 1 slot
  ..@ .Data:List of 3
  .. ..$ :Formal class 'VarLevelSplit' [package "rtables"] with 20 slots
  .. ..$ :Formal class 'VarLevelSplit' [package "rtables"] with 20 slots
  .. ..$ :Formal class 'AnalyzeMultiVars' [package "rtables"] with 17 slots
```

The last print is very informative. We can see from the layout construction that this object is built with 2 `VarLevelSplit`s on the rows and one final `AnalyzeMultiVars`, which is the leaf analysis split that has the final level rows. The second split vector is the following `AnalyzeVarSplit`.

xxx To get multiple split vectors, you need to escape the nesting with `nest = FALSE` or by adding a `split_rows_by` call after an `analyze` call.

```c
# rtables 0.6.2
str(rlyt[[2]], max.level = 5)
Formal class 'SplitVector' [package "rtables"] with 1 slot
  ..@ .Data:List of 1
  .. ..$ :Formal class 'AnalyzeVarSplit' [package "rtables"] with 21 slots
  .. .. .. ..@ analysis_fun           :function (x, ...)
  .. .. .. .. ..- attr(*, "srcref")= 'srcref' int [1:8] 1723 5 1732 5 5 5 4198 4207
  .. .. .. .. .. ..- attr(*, "srcfile")=Classes 'srcfilealias', 'srcfile' <environment: 0x560d8e67b750>
  .. .. .. ..@ default_rowlabel       : chr "Var3 Counts"
  .. .. .. ..@ include_NAs            : logi FALSE
  .. .. .. ..@ var_label_position     : chr "default"
  .. .. .. ..@ payload                : chr "VAR3"
  .. .. .. ..@ name                   : chr "VAR3"
  .. .. .. ..@ split_label            : chr "Var3 Counts"
  .. .. .. ..@ split_format           : NULL
  .. .. .. ..@ split_na_str           : chr NA
  .. .. .. ..@ split_label_position   : chr(0)
  .. .. .. ..@ content_fun            : NULL
  .. .. .. ..@ content_format         : NULL
  .. .. .. ..@ content_na_str         : chr(0)
  .. .. .. ..@ content_var            : chr ""
  .. .. .. ..@ label_children         : logi FALSE
  .. .. .. ..@ extra_args             : list()
  .. .. .. ..@ indent_modifier        : int 0
  .. .. .. ..@ content_indent_modifier: int 0
  .. .. .. ..@ content_extra_args     : list()
  .. .. .. ..@ page_title_prefix      : chr NA
  .. .. .. ..@ child_section_div      : chr NA
```

Continuing in `recursive_applysplit`, this is made up of two main calls: one to `.make_ctab` which makes the content row and calculates the counts if specified, and `.make_split_kids`. This eventually contains `recursive_applysplit` which is applied if the split vector is built of `Split`s that are not `analyze` splits. It being a generic is very handy here to switch between different downstream processes. In our case (`rlyt[[1]]`) we will call the method `getMethod(".make_split_kids", "Split")` twice before getting to the analysis split. There, we have a (xxx) multi-variable split which applies `.make_split_kids` to each of its elements, in turn calling the main `getMethod(".make_split_kids", "VAnalyzeSplit")` which would in turn go to `.make_analyzed_tab`.

There are interesting edge cases here for different split cases, like `split_by_multivars` or when one of the splits has a reference group. In the internal code here, it is called `baseline`. If we follow this variable across the function layers, we will see that where the split (`do_split`) happens (in `getMethod(".make_split_kids", "Split")`) we have a second split for the reference group. This is done to make this available in each row to calculate, for example, differences from the reference group.

Now we move towards `.make_tablerows`, and here analysis functions become key as this is the place where these are applied and analyzed. First, the external `tryCatch` is used to cache errors at a higher level, so as to differentiate the two major blocks. The function parameters here are quite intuitive, with the exception of `spl_context`. This is a fundamental parameter that keeps information about splits so that it can be visible from analysis functions. If you look into this value, you will see that is carried and updated everywhere a split happens, except for columns. Column-related information is added last, when in `gen_onerv`, which is the lowest level where one result value is produced. From `.make_tablerows` we go to `gen_rowvalues`, aside from some row and referential footers handling. `gen_rowvalues` unpacks the `cinfo` object and crosses it with the arriving row split information to generate rows. In particular, `rawvals <- mapply(gen_onerv,` maps the columns to generate a list of values corresponding to a table row. Looking at the final `if` in `gen_onerv` we see `if (!is(val, "RowsVerticalSection"))` and the function `in_rows` is called. We invite the reader to explore what the building blocks of `in_rows` are, and how `.make_tablerows` constructs a data row (`DataRow`) or a content row (`ContentRow`) depending on whether it is called from `.make_ctab` or `.make_analyzed_tab`.

`.make_tablerows` either makes a content table or an "analysis table".
`gen_rowvalues` generates a list of stacks (`RowsVerticalSection`, more than one rows potentially!) for each column.

To add: conceptual part -> calculating things by column and putting them side by side and slicing them by rows and putting it together -> rtables is row dominant.
